---
title: "2c: Local Measures of Spatial Autocorrelation"
author: "Magdalene Chan"
date: 2023-11-24
date-modified: "last-modified"
execute: 
  warning: false
---

Spatial autocorrelation is the term used to describe the presence of systematic spatial variation in a variable. Where there is positive (high) spatial autocorrelation, there is spatial clustering and neighbours are similar. Conversely, where there is negative (low) spatial autocorrelation, checkerboard patterns are observed and neighbours are dissimilar.

Spatial autocorrelation can be used in the development of spatial policy, where one of the main development objectives of governments and planners is to ensure equal distribution of development in the area. Appropriate spatial statistical methods can be applied to discover if developments are evenly distributed geographically -- if there are signs of spatial clustering and if so, where they are located.

## Getting Started

The code chunk below uses p_load() of pacman package to check if the required packages have been installed on the computer. If they are, the packages will be launched.

-   **sf** package is used for importing, managing, and processing geospatial data.
-   **tmap** package is used for thematic mapping.
-   **spdep** package is used to create spatial weights matrix objects.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

The data sets used are:

-   Hunan county boundary layer: a geospatial data set in ESRI shapefile format.
-   Hunan_2012.csv: csv file that contains selected Hunan's local development indicators in 2012.

## Importing Data

### Import shapefile into R

The code chunk below uses the `st_read()` function of **sf** package to import Hunan county boundary shapefile into R as a simple feature data frame called `hunan`.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

There are a total of 88 polygon features and 7 fields in `hunan` simple feature data frame. `hunan` is in **wgs84** GCS.

### Import aspatial data into R

The code chunk below uses the `read_csv()` function of **readr** package to import `Hunan_2012.csv` file into R and save it as a R dataframe called `hunan2012`.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### Performing relational join

`left_join()` of **dplyr** is used to join the geographical data and attribute table using `County` as the common identifier.

```{r}
hunan <- left_join(hunan,hunan2012)%>%
  select(1:4, 7, 15)
```

## Visualising Regional Development Indicator

Prepare a basemap and a choropleth map showing the distribution of GDPPC 2012 by using `qtm()` of **tmap** package.

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```


